<!-- Title can be changed as well as the export format -->
---
title: "Malus_procrustes_analysis_R"
format: html
 editor: visual
---

<!-- Preliminary modules, libraries and functions for the rest of the code to function -->

```{r}
#| label: R Libraries
#| echo: false
#########################
### LOAD IN LIBRARIES ###
#########################

#python libraries that will be used in the python sections
#test to make sure these will load
#library(reticulate)
#py_install("opencv-python")
#py_install("scikit-learn")
#py_install("matplotlib")
#py_install("pandas")
#py_install("numpy")
#py_install("seaborn")

# other r libraries
library(tidyverse) # used for ggplot2
```

```{python}
#| label: Python Libraries
#| echo: false
#######################
### LOAD IN MODULES ###
#######################

import cv2 # to install on mac: pip install opencv-python
from scipy.interpolate import interp1d # for interpolating points
from sklearn.decomposition import PCA # for principal component analysis
from scipy.spatial import procrustes # for Procrustes analysis
from scipy.spatial import ConvexHull # for convex hull
from sklearn.discriminant_analysis import LinearDiscriminantAnalysis # for LDA
from sklearn.metrics import confusion_matrix # for confusion matrix
from os import listdir # for retrieving files from directory
from os.path import isfile, join # for retrieving files from directory
import matplotlib.pyplot as plt # for plotting
import numpy as np # for using arrays
import math # for mathematical operations
import pandas as pd # for using pandas dataframes
import seaborn as sns # for plotting in seaborn
from matplotlib.colors import LogNorm


```

```{python}
#| label: Python Functions
#| echo: false
#################
### FUNCTIONS ###
#################

def angle_between(p1, p2, p3):
    """
    define a function to find the angle between 3 points anti-clockwise in degrees, p2 being the vertex
    inputs: three angle points, as tuples
    output: angle in degrees
    """
    x1, y1 = p1
    x2, y2 = p2
    x3, y3 = p3
    deg1 = (360 + math.degrees(math.atan2(x1 - x2, y1 - y2))) % 360
    deg2 = (360 + math.degrees(math.atan2(x3 - x2, y3 - y2))) % 360
    return deg2 - deg1 if deg1 <= deg2 else 360 - (deg1 - deg2)

def rotate_points(xvals, yvals, degrees):
    """"
    define a function to rotate 2D x and y coordinate points around the origin
    inputs: x and y vals (can take pandas dataframe columns) and the degrees (positive, anticlockwise) to rotate
    outputs: rotated and y vals
    """
    angle_to_move = 90-degrees
    rads = np.deg2rad(angle_to_move)
    
    new_xvals = xvals*np.cos(rads)-yvals*np.sin(rads)
    new_yvals = xvals*np.sin(rads)+yvals*np.cos(rads)
    
    return new_xvals, new_yvals

def interpolation(x, y, number): 
    """
    define a function to return equally spaced, interpolated points for a given polyline
    inputs: arrays of x and y values for a polyline, number of points to interpolate
    ouputs: interpolated points along the polyline, inclusive of start and end points
    """
    distance = np.cumsum(np.sqrt( np.ediff1d(x, to_begin=0)**2 + np.ediff1d(y, to_begin=0)**2 ))
    distance = distance/distance[-1]

    fx, fy = interp1d( distance, x ), interp1d( distance, y )

    alpha = np.linspace(0, 1, number)
    x_regular, y_regular = fx(alpha), fy(alpha)
    
    return x_regular, y_regular

def euclid_dist(x1, y1, x2, y2):
    """
    define a function to return the euclidean distance between two points
    inputs: x and y values of the two points
    output: the eulidean distance
    """
    return np.sqrt((x2-x1)**2 + (y2-y1)**2)

def poly_area(x,y):
    """
    define a function to calculate the area of a polygon using the shoelace algorithm
    inputs: separate numpy arrays of x and y coordinate values
    outputs: the area of the polygon
    """
    return 0.5*np.abs(np.dot(x,np.roll(y,1))-np.dot(y,np.roll(x,1)))

def gpa_mean(leaf_arr, landmark_num, dim_num):
    
    """
    define a function that given an array of landmark data returns the Generalized Procrustes Analysis mean
    inputs: a 3 dimensional array of samples by landmarks by coordinate values, number of landmarks, number of dimensions
    output: an array of the Generalized Procrustes Analysis mean shape
    
    """

    ref_ind = 0 # select arbitrary reference index to calculate procrustes distances to
    ref_shape = leaf_arr[ref_ind, :, :] # select the reference shape

    mean_diff = 10**(-30) # set a distance between means to stop the algorithm

    old_mean = ref_shape # for the first comparison between means, set old_mean to an arbitrary reference shape

    d = 1000000 # set d initially arbitraily high

    while d > mean_diff: # set boolean criterion for Procrustes distance between mean to stop calculations

        arr = np.zeros( ((len(leaf_arr)),landmark_num,dim_num) ) # empty 3D array: # samples, landmarks, coord vals

        for i in range(len(leaf_arr)): # for each leaf shape 

            s1, s2, distance = procrustes(old_mean, leaf_arr[i]) # calculate procrustes adjusted shape to ref for current leaf
            arr[i] = s2 # store procrustes adjusted shape to array

        new_mean = np.mean(arr, axis=(0)) # calculate mean of all shapes adjusted to reference

        s1, s2, d = procrustes(old_mean, new_mean) # calculate procrustes distance of new mean to old mean

        old_mean = new_mean # set the old_mean to the new_mean before beginning another iteration

    return new_mean
```

<!-- Following sections are from the domestic apples. Reads data and does the preliminary image analysis -->

### Domestic Apple Data

#### Metadata Listing

```{python}
#| label: Domestic Apple Metadata
########################
### READ IN METADATA ###
########################

# READ IN CSV FROM CORRECT DIRECTORY
mdata = pd.read_csv('Malus_Domestic_apple_metadata.csv') # read in csv

mdata.head(50) # head data to check
```

```{python}
#| label: Domestic Apple List
#######################################
### MAKE A LIST OF IMAGE FILE NAMES ###
#######################################

# READ IN CSV FROM CORRECT DIRECTORY
data_dir = 'Malus_domestic_JPEGs\\' # set data directory # make sure there is a / at the end of the path

file_names = [f for f in listdir(data_dir) if isfile(join(data_dir, f))] # create a list of file names


file_names.sort() # sort the list of file names

 # check list of file names
file_names[slice(20)]
```

#### Process and Landmarking

-   Read in image in grayscale
-   Select the contour of the largest object (the leaf)
-   Interpolate with a high resolution of pseudo-landmarks
-   Find the base and tip index point on the high resolution contour
-   Reset the base index to zero
-   Interpolate each side with desired number of equidistant pseudo-landmarks
-   Rotate leaves and scale to centimeters
-   Save pseudo-landmarks scaled to centimeters in an array

##### PARAMETERS AND INDEXING:

-   `high_res_pts` is an arbitrarily high number of points to initially interpolate
-   `res` is the desired number of points to interpolate on each side of the leaf
-   The total number of pseudo-landmarks will be `2*res - 1`
-   The base index will be `0`
-   The tip index will be `res-1`
-   The returned leaves in `domestic_cm_arr` are scaled in size to centimeters

```{python}
#| label: Domestic Apple Parameters
######################
### SET PARAMETERS ###
######################

# the number of equidistant points to create
# an initial high resolution outline of the leaf
high_res_pts = 1000 

# the ultimate number of equidistant points on each side of the leaf
# (-1 for the tip)
# the leaf will have res*2-1 pseudo-landmarks
#################
#################
#################
res = 20 ########
#################
#################
#################

# an array to store pseudo-landmarks
domestic_cm_arr = np.zeros((len(mdata),(res*2)-1,2))

# for each leaf . . .
for lf in range(len(mdata)):

    ###############################
    ### READ IN GRAYSCALE IMAGE ###
    ###############################

    curr_image = mdata["file"][lf] # select the current image
    print(lf, curr_image) # print each leaf in case there are problems later
    
    # read in image
    # convert to grayscale
    # invert the binary
    img = cv2.bitwise_not(cv2.cvtColor(cv2.imread(data_dir + curr_image),cv2.COLOR_BGR2GRAY))

    # find contours of binary objects
    contours, hierarchy = cv2.findContours(img,  
        cv2.RETR_EXTERNAL, cv2.CHAIN_APPROX_SIMPLE)

    ##############################
    ### SELECT LARGEST CONTOUR ###
    ##############################

    # ideally there is only one leaf in the image
    # in the case there are smaller objects
    # this code selects the largest object (the leaf)
    # if there is one and only one object in the image
    # then the following code is not necessary

    x_conts = [] # list of lists of contour x vals
    y_conts = [] # list of lists of contour y vals
    areas_conts = [] # list of bounding box areas of contours
    for c in contours: # for each contour
        x_vals = [] # store x vals for current contour 
        y_vals = [] # store y vals for current contour
        for i in range(len(c)): # for each point in current contour
            x_vals.append(c[i][0][0]) # isolate x val
            y_vals.append(c[i][0][1]) # isolate y val
        area = (max(x_vals) - min(x_vals))*(max(y_vals) - min(y_vals)) # calculate bounding box area of contour
        x_conts.append(x_vals) # append the current contour x vals
        y_conts.append(y_vals) # append the current contour y vals
        areas_conts.append(area) # append the current contour bounding box areas

    area_inds = np.flip(np.argsort(areas_conts)) # get indices to sort contours by area
    sorted_x_conts = np.array(x_conts, dtype=object)[area_inds][0:] # areas sorted largest to smallest, x vals
    sorted_y_conts = np.array(y_conts, dtype=object)[area_inds][0:] # areas sorted largest to smallest, y vals

    ################################################
    ### INTERPOLATE HIGH RES NUMBER OF LANDMARKS ###
    ################################################

    # convert the leaf to high resolution number of landmarks
    # using high_res_pt value
    # need to convert arrays of pixel int to floats first
    high_res_x, high_res_y = interpolation(np.array(sorted_x_conts[0], dtype=np.float32), 
                                           np.array(sorted_y_conts[0], dtype=np.float32), high_res_pts)

    ###############################
    ### FIND BASE AND TIP INDEX ###
    ###############################

    # get the base and tip landmark point values
    base_pt = np.array((mdata["base_x"][lf], mdata["base_y"][lf]))
    tip_pt = np.array((mdata["tip_x"][lf], mdata["tip_y"][lf]))

    base_dists = [] # store distance of each high res point to base
    tip_dists = [] # store distance of each high res point to tip

    for pt in range(len(high_res_x)): # for each of the high resolution points

        # euclidean distance of the current point from the base and tip landmark
        ed_base = euclid_dist(base_pt[0], base_pt[1], high_res_x[pt], high_res_y[pt])
        ed_tip = euclid_dist(tip_pt[0], tip_pt[1], high_res_x[pt], high_res_y[pt])

        # store distance of current point from base/tip
        base_dists.append(ed_base)
        tip_dists.append(ed_tip)

    # get index of base and tip points
    base_ind = np.argmin(base_dists)
    tip_ind = np.argmin(tip_dists)

    ################################
    ### RESET BASE INDEX TO ZERO ###
    ################################

    # reset base index position to zero
    high_res_x = np.concatenate((high_res_x[base_ind:],high_res_x[:base_ind]))
    high_res_y = np.concatenate((high_res_y[base_ind:],high_res_y[:base_ind]))

    # recalculate indices with new indexing
    tip_ind = tip_ind-base_ind # note: negative index if tip_ind<base_ind
    base_ind = base_ind-base_ind

    # create single array for leaf coordinates
    lf_contour = np.column_stack((high_res_x, high_res_y))

    ##############################################################
    ### INTERPOLATE EACH SIDE WITH DESIRED NUMBER OF LANDMARKS ###
    ##############################################################

    # interpolate at desired resolution the left and right sides of the leaf
    left_inter_x, left_inter_y = interpolation(lf_contour[base_ind:tip_ind+1,0],lf_contour[base_ind:tip_ind+1,1],res)
    right_inter_x, right_inter_y = interpolation(lf_contour[tip_ind:,0],lf_contour[tip_ind:,1],res)

    # the start of the right side and end of the left side
    # both contain the tip landmark
    # delete the last point on the left side
    left_inter_x = np.delete(left_inter_x, -1)
    left_inter_y = np.delete(left_inter_y, -1)

    # BASE OF LEAF IS INDEX 0
    # TIP INDEX IS RES-1 IF BOTH LEFT & RIGHT POINTS
    # TOTAL PSEUDOLANDMARKS IS 2*RES-1
    lf_pts_left = np.column_stack((left_inter_x, left_inter_y))
    lf_pts_right = np.column_stack((right_inter_x, right_inter_y))
    lf_pts = np.row_stack((lf_pts_left, lf_pts_right))

    ##########################################################
    ### ROTATE LEAVES UPWARD AND SCALE SIZE TO CENTIMETERS ###
    ##########################################################

    tip_point = lf_pts[res-1,:] # get tip point
    base_point = lf_pts[0,:] # get base point

    # calculate angle between tip. base, and an arbitrary reference
    ang = angle_between(tip_point, base_point, (base_point[0]+1,base_point[1]) )

    # rotate points upwards
    rot_x, rot_y = rotate_points(lf_pts[:,0], lf_pts[:,1], ang) 
    rot_pts = np.column_stack((rot_x, rot_y))
    
    # calculate leaf area in pixels^2
    lf_area_px2 = poly_area(rot_pts[:,0], rot_pts[:,1])
    
    # get px_cm
    px_cm = np.sqrt(lf_area_px2/mdata["area"][lf])

    # scale leaf into cm
    cm_lf = rot_pts/(px_cm)
    
    # store the leaf scaled into cm into the cm_arr
    domestic_cm_arr[lf,:,:] = cm_lf
```

<!-- Following sections are from the wild apples. Reads data and does the preliminary image analysis -->

### Wild Apple Data

#### Metadata Listing

```{python}
#| label: Wild Apple Metadata
########################
### READ IN METADATA ###
########################

# CHANGE TO CORRECT DIRECTORY
mdata = pd.read_csv("C:\\Users\\Lachlann\\OneDrive - Acadia University\\Documents\\Apple_metadata_thesis\\Metadata\\Malus_wild_metadata.csv") # read in csv

mdata.head(50) # head data to check
```

```{python}
#| label: Wild Apple List
#######################################
### MAKE A LIST OF IMAGE FILE NAMES ###
#######################################

# READ IN CSV FROM CORRECT DIRECTORY
data_dir = ("C:\\Users\\Lachlann\\OneDrive - Acadia University\\Documents\\Apple_metadata_thesis\\Malus_wild_binary_files\\")
#data directory

file_names = [f for f in listdir(data_dir) if isfile(join(data_dir, f))] # create a list of file names

#file_names.remove('.DS_Store') # remove .DS_Store file

file_names.sort() # sort the list of file names

file_names[slice(50)] # check list of file names
```

#### Process and Landmarking

-   Read in image in grayscale
-   Select the contour of the largest object (the leaf)
-   Interpolate with a high resolution of pseudo-landmarks
-   Find the base and tip index point on the high resolution contour
-   Reset the base index to zero
-   Interpolate each side with desired number of equidistant pseudo-landmarks
-   Rotate leaves and scale to centimeters
-   Save pseudo-landmarks scaled to centimeters in an array

##### PARAMATERS AND INDEXING:

-   `high_res_pts` is an arbitrarily high number of points to initially interpolate
-   `res` is the desired number of points to interpolate on each side of the leaf
-   The total number of pseudo-landmarks will be `2*res - 1`
-   The base index will be `0`
-   The tip index will be `res-1`
-   The returned leaves in `wild_cm_arr` are scaled in size to centimeters

```{python}
#| label: Wild Apple Parameters
######################
### SET PARAMETERS ###
######################

# the number of equidistant points to create
# an initial high resolution outline of the leaf
high_res_pts = 1000 

# the ultimate number of equidistant points on each side of the leaf
# (-1 for the tip)
# the leaf will have res*2-1 pseudo-landmarks
#################
#################
#################
res = 20 ########
#################
#################
#################

# an array to store pseudo-landmarks
wild_cm_arr = np.zeros((len(mdata),(res*2)-1,2))

# for each leaf . . .
for lf in range(len(mdata)):

    ###############################
    ### READ IN GRAYSCALE IMAGE ###
    ###############################

    curr_image = mdata["file"][lf] # select the current image
    print(lf, curr_image) # print each leaf in case there are problems later

    # read in image
    # convert to grayscale
    # invert the binary
    img = cv2.bitwise_not(cv2.cvtColor(cv2.imread(data_dir + curr_image),cv2.COLOR_BGR2GRAY))

    # find contours of binary objects
    contours, hierarchy = cv2.findContours(img,  
        cv2.RETR_EXTERNAL, cv2.CHAIN_APPROX_SIMPLE)

    ##############################
    ### SELECT LARGEST CONTOUR ###
    ##############################

    # ideally there is only one leaf in the image
    # in the case there are smaller objects
    # this code selects the largest object (the leaf)
    # if there is one and only one object in the image
    # then the following code is not necessary

    x_conts = [] # list of lists of contour x vals
    y_conts = [] # list of lists of contour y vals
    areas_conts = [] # list of bounding box areas of contours
    for c in contours: # for each contour
        x_vals = [] # store x vals for current contour 
        y_vals = [] # store y vals for current contour
        for i in range(len(c)): # for each point in current contour
            x_vals.append(c[i][0][0]) # isolate x val
            y_vals.append(c[i][0][1]) # isolate y val
        area = (max(x_vals) - min(x_vals))*(max(y_vals) - min(y_vals)) # calculate bounding box area of contour
        x_conts.append(x_vals) # append the current contour x vals
        y_conts.append(y_vals) # append the current contour y vals
        areas_conts.append(area) # append the current contour bounding box areas

    area_inds = np.flip(np.argsort(areas_conts)) # get indices to sort contours by area
    sorted_x_conts = np.array(x_conts, dtype=object)[area_inds][0:] # areas sorted largest to smallest, x vals
    sorted_y_conts = np.array(y_conts, dtype=object)[area_inds][0:] # areas sorted largest to smallest, y vals

    ################################################
    ### INTERPOLATE HIGH RES NUMBER OF LANDMARKS ###
    ################################################

    # convert the leaf to high resolution number of landmarks
    # using high_res_pt value
    # need to convert arrays of pixel int to floats first
    high_res_x, high_res_y = interpolation(np.array(sorted_x_conts[0], dtype=np.float32), 
                                           np.array(sorted_y_conts[0], dtype=np.float32), high_res_pts)

    ###############################
    ### FIND BASE AND TIP INDEX ###
    ###############################

    # get the base and tip landmark point values
    base_pt = np.array((mdata["base_x"][lf], mdata["base_y"][lf]))
    tip_pt = np.array((mdata["tip_x"][lf], mdata["tip_y"][lf]))

    base_dists = [] # store distance of each high res point to base
    tip_dists = [] # store distance of each high res point to tip

    for pt in range(len(high_res_x)): # for each of the high resolution points

        # euclidean distance of the current point from the base and tip landmark
        ed_base = euclid_dist(base_pt[0], base_pt[1], high_res_x[pt], high_res_y[pt])
        ed_tip = euclid_dist(tip_pt[0], tip_pt[1], high_res_x[pt], high_res_y[pt])

        # store distance of current point from base/tip
        base_dists.append(ed_base)
        tip_dists.append(ed_tip)

    # get index of base and tip points
    base_ind = np.argmin(base_dists)
    tip_ind = np.argmin(tip_dists)

    ################################
    ### RESET BASE INDEX TO ZERO ###
    ################################

    # reset base index position to zero
    high_res_x = np.concatenate((high_res_x[base_ind:],high_res_x[:base_ind]))
    high_res_y = np.concatenate((high_res_y[base_ind:],high_res_y[:base_ind]))

    # recalculate indices with new indexing
    tip_ind = tip_ind-base_ind # note: negative index if tip_ind<base_ind
    base_ind = base_ind-base_ind

    # create single array for leaf coordinates
    lf_contour = np.column_stack((high_res_x, high_res_y))

    ##############################################################
    ### INTERPOLATE EACH SIDE WITH DESIRED NUMBER OF LANDMARKS ###
    ##############################################################

    # interpolate at desired resolution the left and right sides of the leaf
    left_inter_x, left_inter_y = interpolation(lf_contour[base_ind:tip_ind+1,0],lf_contour[base_ind:tip_ind+1,1],res)
    right_inter_x, right_inter_y = interpolation(lf_contour[tip_ind:,0],lf_contour[tip_ind:,1],res)

    # the start of the right side and end of the left side
    # both contain the tip landmark
    # delete the last point on the left side
    left_inter_x = np.delete(left_inter_x, -1)
    left_inter_y = np.delete(left_inter_y, -1)

    # BASE OF LEAF IS INDEX 0
    # TIP INDEX IS RES-1 IF BOTH LEFT & RIGHT POINTS
    # TOTAL PSEUDOLANDMARKS IS 2*RES-1
    lf_pts_left = np.column_stack((left_inter_x, left_inter_y))
    lf_pts_right = np.column_stack((right_inter_x, right_inter_y))
    lf_pts = np.row_stack((lf_pts_left, lf_pts_right))

    ##########################################################
    ### ROTATE LEAVES UPWARD AND SCALE SIZE TO CENTIMETERS ###
    ##########################################################

    tip_point = lf_pts[res-1,:] # get tip point
    base_point = lf_pts[0,:] # get base point

    # calculate angle between tip. base, and an arbitrary reference
    ang = angle_between(tip_point, base_point, (base_point[0]+1,base_point[1]) )

    # rotate points upwards
    rot_x, rot_y = rotate_points(lf_pts[:,0], lf_pts[:,1], ang) 
    rot_pts = np.column_stack((rot_x, rot_y))

    # scale leaf into cm
    cm_lf = rot_pts/(mdata["px-cm"][lf])
    
    # store the leaf scaled into cm into the cm_arr
    wild_cm_arr[lf,:,:] = cm_lf
```

<!-- Merging sections of the data together -->

#### Combining Domestic and Wild

```{python}
#| label: Combining Array
# stack domesticated and wild arrays into one new array

cm_arr = np.row_stack(( domestic_cm_arr,wild_cm_arr))


```

```{python}
xset= (cm_arr[:,:,0])
yset= (cm_arr[:,:,1])
lngset= np.concatenate((xset,yset), axis=1)
newcolumns=[f'x{i+1}' for i in range(39)] + [f'y{i+1}' for i in range(39)]


```
#### Combining Datasets

```{python}
#| label: Combining Datasets
# read in each dataset

# READ IN CSV FROM CORRECT DIRECTORY
domesticated_data = pd.read_csv("Malus_Domestic_apple_metadata.csv") # read in domesticated csv


# READ IN CSV FROM CORRECT DIRECTORY
wild_data = pd.read_csv("Malus_wild_metadata.csv") # read in wild csv

# stack domesticated and wild genotypes into one new dataframe
combined_df = pd.concat([domesticated_data, wild_data], ignore_index=True)
#combined_df.to_csv("", index=False)
len(combined_df)

```

<!-- The following sections of code are from the analysis -->

### Analyzing Leaf Dimensions


Using placed pseudo-landmarks representing leaves that are rotated upwards and scaled in centimeters from `cm_arr`, calculate the following:

-   `width`: difference in centimeters between minimum and maximum x values in an oriented leaf
-   `length`: difference in centimeters between minimum and maximum y values in an oriented leaf
-   `area`: area of the leaf in centimeters squared
-   `solidity`: the ratio of area to convex hull area
-   `asymmetry`: the Procrustes distance between the superimposed left and right sides of a leaf outline. Lower values are more symmetric. Higher values are more asymmetric.

Data is stored in the `combined_df` dataframe.

```{python}
#| label: Analyzing Leaf Dimensions
# lists to store variables
width_list = []
length_list = []
area_list = []
solidity_list = []
# for each leaf . . .
for lf in range(len(cm_arr)):
    
    # for calculating dimensions, we need non-scaled leaves in centimeters
    curr_lf = cm_arr[lf,:,:] # select current leaf
    
    ############################
    ### CALCULATE DIMENSIONS ###
    ############################
    
    width = np.max(curr_lf[:,0])-np.min(curr_lf[:,0]) # calculate width
    length = np.max(curr_lf[:,1])-np.min(curr_lf[:,1]) # calculate length
    area = poly_area(curr_lf[:,0],curr_lf[:,1]) # calculate area
    
    ##########################
    ### CALCULATE SOLIDITY ### potentially show difference in lobing
    ##########################
    
    hull = ConvexHull(curr_lf) # calculate convex hull of current leaf
    vertices = hull.vertices # isolate vertex indices of convex hull
    convex_area = poly_area(curr_lf[vertices,0], curr_lf[vertices,1]) # calculate convex area
    solidity = area / convex_area # calculate solidity
    

    # store data in lists
    width_list.append(width)
    length_list.append(length)
    area_list.append(area)
    solidity_list.append(solidity)
    

# add data to the combined_df dataframe
combined_df["width"] = width_list
combined_df["length"] = length_list
combined_df["area"] = area_list
combined_df["solidity"] = solidity_list

# To analyze leaf size/allometry
# calculate leaf area of each leaf in cm
# and replace 'area' column values in combined_df

leaf_areas_cm = []

for lf in range(len(cm_arr)):
    
    leaf_areas_cm.append(poly_area(cm_arr[lf,:,0], cm_arr[lf,:,1]))

combined_df["area"] = leaf_areas_cm

#combined_df.to_csv("", index=False)

```

<!-- PLEASE INSERT PLOT BELOW -->
<!-- NOTE: will need to convert datasets that will be used from python to r. This can be done by using the py_to_r function from the reticulate library-->

### Figure 4
```{r}
#Fig 4. PC1 and width, length, and area
# Create the three ggplot objects
#Apple_metadata<- read.csv("Apple_metadata.csv")

plot4 <- ggplot(Apple_metadata, aes(x=PC1, y=log(width))) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank()   
  )+  geom_smooth(method=lm, color="red", se=FALSE) +
      scale_x_continuous(breaks = pretty(AspectMeta$PC1, n = 5)) +
  geom_smooth(method=lm, color="red", se=FALSE) +
   labs(
    x = "PC1",  
    y = "ln(width, cm)"   
  )

plot5 <- ggplot(Apple_metadata, aes(x=PC1, y=log(length))) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank()   
  )+  geom_smooth(method=lm, color="red", se=FALSE) +
      scale_x_continuous(breaks = pretty(AspectMeta$PC1, n = 5)) +
   labs(
    x = "PC1",  
    y = "ln(length, cm)"    
  )

plot6 <- ggplot(Apple_metadata, aes(x=PC1, y=log(area))) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()  
  )+  geom_smooth(method=lm, color="red", se=FALSE) +
    scale_x_continuous(breaks = pretty(AspectMeta$PC1, n = 5)) +
   labs(
    x = "PC1", 
    y = "ln(area, cm²)"    
  )

# Combine the plots into a 1x3 grid
#grid.arrange(plot4, plot5, plot6, nrow = 1) 
library(gtable)
g4 <- ggplotGrob(plot4)
g5 <- ggplotGrob(plot5)
g6 <- ggplotGrob(plot6)
g <- rbind(g4, g5,g6, size = "first")
g$widths <- unit.pmax(g4$widths, g5$widths, g6$widths)
grid.newpage()
pdf("figure4.pdf", width=4, height=6)
grid.draw(g)
dev.off()
###Additional plot edits were made in Affinity
```

###Figure S1

```{r}
library(ggplot2)
library(GGally)
library(dplyr)
#Apple_metadata<- read.csv("Apple_metadata.csv")

#Label Fig S1. Log of width/area, length/area, length/width
# Create the three ggplot objects
plot1 <- ggplot(Apple_metadata, aes(x=log(length), y=log(width))) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank()   
  )+
  geom_smooth(method=lm, color="red", se=FALSE) +
   labs(
    x = "ln(length, cm)",  
    y = "ln(width, cm)"    
  )
  

plot2 <- ggplot(Apple_metadata, aes(x=log(length), y=log(area))) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank()   
  )+
  geom_smooth(method=lm, color="red", se=FALSE) +
   labs(
    x = "ln(length, cm)",  
    y = "ln(area, cm²)"    
  )
  

plot3 <- ggplot(Apple_metadata, aes(x=log(width), y=log(area))) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()   
  )+
  geom_smooth(method=lm, color="red", se=FALSE) +
   labs(
    x = "ln(width, cm)",  
    y = "ln(area, cm²)"   
  )

# Combine the plots into a 1x3 grid
#grid.arrange(plot1, plot2, plot3, ncol=2)
library(gtable)
g1 <- ggplotGrob(plot1)
g2 <- ggplotGrob(plot2)
g3 <- ggplotGrob(plot3)
g <- rbind(g1, g2,g3, size = "first")
g$widths <- unit.pmax(g1$widths, g2$widths, g3$widths)
grid.newpage()
pdf("Supplementalfigure1.pdf", width=6, height=8)
grid.draw(g)
dev.off()

```

### Calculate PC

```{python}
<!-- The following sections are the Procrustes Analysis -->

### Procrustes Analysis

Perform a Procrustes analysis to translate, scale, and rotate leaf shapes

#-   Select number of pseudo-landmarks and dimensions
#-   Calculate the GPA mean leaf shape using the `gpa_mean` function
#-   Align all leaves to the GPA mean
#-   Store Procrustes super-imposed leaves in an array, `proc_arr`
#-   Calculate a PCA for all possible axes and their variance (the number of leaves)
#-   Calculate a PCA for just the axes needed for reconstruction of eigenleaves for morphospace 


#| label: GPA Mean
landmark_num = (res*2)-1 # select number of landmarks
dim_num = 2 # select number of coordinate value dimensions

##########################
### CALCULATE GPA MEAN ###
##########################

mean_shape = gpa_mean(cm_arr, landmark_num, dim_num)

################################
### ALIGN LEAVES TO GPA MEAN ###
################################

# array to store Procrustes aligned shapes
proc_arr = np.zeros(np.shape(cm_arr)) 

for i in range(len(cm_arr)):
    s1, s2, distance = procrustes(mean_shape,cm_arr[i,:,:]) # calculate procrustes adjusted shape to ref for current leaf
    proc_arr[i] = s2 # store procrustes adjusted shape to array

###############################################################

#| label: Calculate PC
#################################################
### FIRST, CALCULATE PERCENT VARIANCE ALL PCs ###
#################################################

######
PC_NUMBER = 78 # PC number = not number of leaves, but features
#######

# use the reshape function to flatten to 2D
flat_arr = proc_arr.reshape(np.shape(proc_arr)[0], 
                                 np.shape(proc_arr)[1]*np.shape(proc_arr)[2]) 

pca_all = PCA(n_components=PC_NUMBER) 
PCs_all = pca_all.fit_transform(flat_arr) # fit a PCA for all data

# print out explained variance for each PC
print("PC: " + "var, " + "overall ") 
for i in range(len(pca_all.explained_variance_ratio_)):
    print("PC" + str(i+1) + ": " + str(round(pca_all.explained_variance_ratio_[i]*100,1)) + 
          "%, " + str(round(pca_all.explained_variance_ratio_.cumsum()[i]*100,1)) + "%"  )
```

```{python}
#| label: Calculate Number of PCs
#################################################
### NEXT, CALCULATE THE DESIRED NUMBER OF PCs ###
#################################################

######
PC_NUMBER = 6 # PC number = 3, because PC 2 is asymmetry and we want to create morphospace for PC = 1 and 3
#######

pca = PCA(n_components=PC_NUMBER) 
PCs = pca.fit_transform(flat_arr) # fit a PCA for only desired PCs

# print out explained variance for each PC
print("PC: " + "var, " + "overall ") 
for i in range(len(pca.explained_variance_ratio_)):
    print("PC" + str(i+1) + ": " + str(round(pca.explained_variance_ratio_[i]*100,1)) + 
          "%, " + str(round(pca.explained_variance_ratio_.cumsum()[i]*100,1)) + "%"  )
    
# add PCs to dataframe for plotting
combined_df["PC1"] = PCs[:,0]
combined_df["PC2"] = PCs[:,1]
combined_df["PC3"] = PCs[:,2]
combined_df["PC4"] = PCs[:,3]
combined_df["PC5"] = PCs[:,4]
combined_df["PC6"] = PCs[:,5]

#combined_df.to_csv("", index=False)
```


###Table S2


```{r}
#Create correlation table
library(tidyverse)
#Apple_metadata<- read.csv("Apple_metadata.csv")


cor_table <- Apple_metadata %>%
  mutate(
    'ln(area)' = log(area),
    'ln(width)' = log(width),
    'ln(length)' = log(length)
  ) %>%
  dplyr::select('ln(area)', 'ln(width)', 'ln(length)', aspect, solidity, PC1, PC2, PC3, PC4, PC5, PC6)
cor_table <- dplyr::select(Apple_metadata,area, width,length,aspect, solidity, PC1, PC2, PC3, PC4, PC5, PC6) 
pairwise_comp <- matrix(, ncol = ncol(cor_table), nrow = ncol(cor_table))
rownames(pairwise_comp) <- colnames(cor_table)
colnames(pairwise_comp) <- colnames(cor_table)

pairwise_comp_r <- matrix(, ncol = ncol(cor_table), nrow = ncol(cor_table))
rownames(pairwise_comp_r) <- colnames(cor_table)
colnames(pairwise_comp_r) <- colnames(cor_table)


for (i in 1:ncol(cor_table)) {
  name_x = colnames(cor_table)[i]
  for (j in 1:ncol(cor_table)) {
    name_y = colnames(cor_table)[j]
    pairwise_comp[j, i] = cor.test(cor_table[,i], cor_table[,j], method = "spearman")$p.value
    pairwise_comp_r[j, i] = cor.test(cor_table[,i], cor_table[,j], method = "spearman")$estimate
  }
}

correlation_data <- data.frame(
  row = rep(rownames(pairwise_comp_r), times = ncol(pairwise_comp_r)),
  col = rep(colnames(pairwise_comp_r), each = ncol(pairwise_comp_r)),
  correlation = as.vector(pairwise_comp_r),
  pval = as.vector(pairwise_comp))

correlation_data <- correlation_data %>% 
  subset(!is.na(pval)) %>% 
  subset(row != col)

correlation_data <- correlation_data %>% 
  rename("trait 1" = row, "trait 2" = col)
#write.csv(correlation_data, "examineThesis_cors.csv")
# Bonferroni Correction ---------------------------------------------------
pairwise_comp[lower.tri(pairwise_comp)] = NA

pairwise_comp[upper.tri(pairwise_comp)] = p.adjust(pairwise_comp[upper.tri(pairwise_comp)], method = "bonferroni")

#how many correlations
length(pairwise_comp[upper.tri(pairwise_comp)] )


#how many are significant?
table(pairwise_comp[upper.tri(pairwise_comp)] <0.05)
table(correlation_data$pval <0.05)

correlation_data <- data.frame(
  row = rep(rownames(pairwise_comp_r), times = ncol(pairwise_comp_r)),
  col = rep(colnames(pairwise_comp_r), each = ncol(pairwise_comp_r)),
  correlation = as.vector(pairwise_comp_r),
  pval = as.vector(pairwise_comp))

correlation_data <- correlation_data %>% 
  subset(!is.na(pval)) %>% 
  subset(row != col)

correlation_data <- correlation_data %>% 
  rename("trait 1" = row, "trait 2" = col)
#write.csv(correlation_data, "0316tables2.csv")


```

### Figure S2: Eigen Representation of PC

```{python}
#| label: Figure S2: Eigen Representation of PC6 
###################################################
### CREATE EIGEN REPRESENTATIONS OF FIRST 6 PCs ###
###################################################

# calculate standard deviations for each PC
PC1_std = combined_df["PC1"].std()
PC2_std = combined_df["PC2"].std()
PC3_std = combined_df["PC3"].std()
PC4_std = combined_df["PC4"].std()
PC5_std = combined_df["PC5"].std()
PC6_std = combined_df["PC6"].std()

# create list of lists of 6 PC values to reconstruct

PC_vals = [[-2*PC1_std,0,0,0,0,0],
           [-1*PC1_std,0,0,0,0,0],
           [0*PC1_std,0,0,0,0,0],
           [1*PC1_std,0,0,0,0,0],
           [2*PC1_std,0,0,0,0,0],
           [0,-2*PC2_std,0,0,0,0],
           [0,-1*PC2_std,0,0,0,0],
           [0,0*PC2_std,0,0,0,0],
           [0,1*PC2_std,0,0,0,0],
           [0,2*PC2_std,0,0,0,0],
           [0,0,-1*PC3_std,0,0,0],
           [0,0,-2*PC3_std,0,0,0],
           [0,0,0*PC3_std,0,0,0],
           [0,0,1*PC3_std,0,0,0],
           [0,0,2*PC3_std,0,0,0],
           [0,0,0,-2*PC4_std, 0,0],
           [0,0,0,-1*PC4_std, 0,0],
           [0,0,0,0*PC4_std, 0,0],
           [0,0,0,1*PC4_std, 0,0],
           [0,0,0,2*PC4_std, 0,0],
           [0,0,0,0,-2*PC5_std,0],
           [0,0,0,0,-1*PC5_std,0],
           [0,0,0,0,0*PC5_std,0],
           [0,0,0,0,1*PC5_std,0],
           [0,0,0,0,2*PC5_std,0],
           [0,0,0,0,0,-2*PC6_std],
           [0,0,0,0,0,-1*PC6_std],
           [0,0,0,0,0,0*PC6_std],
           [0,0,0,0,0,1*PC6_std],
           [0,0,0,0,0,2*PC6_std]]
          
           

plt.figure(figsize=(5,8))
counter = 1

for i in range(len(PC_vals)):
    
    # create inverse leaf
    inv_leaf = pca.inverse_transform(np.array(PC_vals[i]).reshape(1, -1))
    inv_x = inv_leaf[0, 0::2] # select just inverse x vals
    inv_y = inv_leaf[0, 1::2] # select just inverse y vals
    
    # plot inverse leaf
    plt.subplot(8,5,counter)
    plt.fill(inv_x, inv_y, c="lightgray")
    plt.plot(inv_x, inv_y, c="gray")
    plt.gca().set_aspect("equal")
    plt.axis("off")
    
    
    counter += 1

plt.tight_layout()
plt.show()
#plt.savefig("")
#Additional edits to the figure were made in Affinity
```
### Figure 2

```{python}
#####Label: Mean shape by species (Figure 2)#####
# Compute mean leaf shapes
mean_shapes_by_species = {}
species_list = np.unique(combined_df["genotype"])

for sp in species_list:
    inds = np.where(combined_df["genotype"] == sp)[0]
    if len(inds) == 0:
        continue  
    mean_shapes_by_species[sp] = proc_arr[inds].mean(axis=0)

# Plot
species = list(mean_shapes_by_species.keys())
n = len(species)
cols = 6
rows = int(np.ceil(n / cols))

plt.figure(figsize=(cols*3, rows*3)) 

for i, sp in enumerate(species):
    plt.subplot(rows, cols, i+1)
    shape = mean_shapes_by_species[sp]
    plt.plot(shape[:,0], shape[:,1], 'k-')
    plt.gca().set_aspect('equal')
    plt.axis('off')
    plt.title(sp, fontsize=7)

plt.suptitle("Mean Procrustes shape per species", y=1.00)
plt.tight_layout()
plt.show()

#plt.savefig("Figure2.pdf")
##Additonal edits to the figure were carried out in affinity
```


### Morphospace and Linear Discriminant Analysis (By Genotype)

Visualize a morphospace and classify leaves by the factor of genotype

-   Plot a morphospace using the inverse transform of the PCA
-   Perform a Linear Discriminant Analysis
-   Visualize LDA results as a confusion matrix
-   Plot out LDA scores
-   Note: the column names to plot the Linear Discriminants are manually renamed

### Figure 3

<!-- Visualize morpho-space's with classification by genotype and continent -->

-   Plot a morphospace using the inverse transform of the PCA

```{python}
#| label: Morphospace aspect (PC1 and PC3)
##########################
### CREATE MORPHOSPACE ###
##########################
from install colour
##########################
### CREATE MORPHOSPACE ###
##########################
Newdata=pd.read_csv("Tree_level_metadata.csv")

# set plot parameters

plot_length= 20 # plot length in inches
plot_width= 20 # plot length in inches
numPC1 = 10 # set number of PC1 intervals
numPC3 = 5 # set number of PC3 intervals
hue = "aspect" # select the factor to color by
s = 0.125 # set the scale of the eigenleaves
lf_col = "lightgray" # color of inverse eigenleaf
lf_alpha = 0.5 # alpha of inverse eigenleaf
pt_size = 20 # size of data points
pt_linewidth = 0 # lw of data points, set to 0 for no edges
pt_alpha = 0.6 # alpha of the data points
ax_label_fs = 12 # font size of the x and y axis titles
ax_tick_fs = 8 # font size of the axis ticks
face_col = "white" # color of the plot background
grid_alpha = 0.5 # set the alpha of the grid


plt.figure(figsize=(plot_length, plot_width))

# note that PC2 is asymmetry and always = 0
PC1_vals = np.linspace( np.min(PCs[:,0]), np.max(PCs[:,0]), numPC1 ) # create PC intervals
PC3_vals = np.linspace( np.min(PCs[:,2]), np.max(PCs[:,2]), numPC3 )

for i in PC1_vals: # for each PC1 interval
    for j in PC3_vals: # for each PC2 interval
        
        pc1_val = i # select the current PC1 val
        pc3_val = j # select the current PC3 val

        # calculate the inverse eigenleaf
        inv_leaf = pca.inverse_transform(np.array([pc1_val,0,pc3_val,0,0,0]).reshape(1, -1))
        inv_x = inv_leaf[0,0::2] # select just inverse x vals
        inv_y = inv_leaf[0,1::2] # select just inverse y vals
        
        # plot the inverse eigenleaf
        plt.fill(inv_x*s+pc1_val, inv_y*s+pc3_val, c=lf_col, alpha=lf_alpha)

# Normalize the color scale based on the aspect ratio limits
norm = plt.Normalize(vmin=0.25, vmax=0.85)  # normalization range for aspect ratio

# Create the scatter plot with the normalized colormap
aspectscatter = sns.scatterplot(
    data=Newdata,
    x="PC1",
    y="PC3",
    hue=hue,
    s=pt_size,
    linewidth=pt_linewidth,
    alpha=pt_alpha,
    palette="viridis",
    legend=False,  # Disable the default legend to avoid overlap with color bar
    hue_norm=norm  # Explicitly set normalization here
)

# Add a colorbar with correct normalization
sm = plt.cm.ScalarMappable(cmap="viridis", norm=norm)
sm.set_array([])  # ScalarMappable requires an empty array here
cbar = plt.colorbar(sm, ax=plt.gca())  # Add colorbar with the correct normalization
cbar.set_ticks(np.linspace(norm.vmin, norm.vmax, 7)) # adding the min and max values, plus evenly spaced out other values
cbar.set_label("Aspect")  # Label the color bar

# Set axis labels and titles
xlab = "PC1 (49.0%) " 
ylab = "PC3 (11.2%) "
plt.xlabel(xlab, fontsize=ax_label_fs)
plt.ylabel(ylab, fontsize=ax_label_fs)
plt.xticks(fontsize=ax_tick_fs)
plt.yticks(fontsize=ax_tick_fs)
plt.gca().set_aspect("equal")
plt.gca().set_facecolor(face_col)
plt.grid(alpha=grid_alpha)
plt.gca().set_axisbelow(True)


plt.show()

#plt.savefig("NewestMorphoAspect.pdf")
##Further plot edits made in Affinity
```

```{r}
#Fig 3 a and b. PC1/Aspect and PC3/Solidity
#Apple_metadata<- read.csv("Apple_metadata.csv")

ggplot(AspectMeta, aes(x=PC1, y=aspect)) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()   
  )+  geom_smooth(method=lm, color="red", se=FALSE) +
 labs(
    x = "PC1", 
    y = "Aspect Ratio"
  )
#ggsave("", plot=last_plot())
ggplot(AspectMeta, aes(x=PC3, y=solidity)) + 
  geom_point() +
theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()  
  )+
  geom_smooth(method=lm, color="red", se=FALSE) +
labs(
    x = "PC3",  
    y = "Solidity"   
  )

#ggsave("", plot=last_plot())

## Morpho-space and ggplots combined in Affinity
```


###Figure 5


```{python}
#| label: Continent Morphospace
##########################
### CREATE MORPHOSPACE ###
##########################
cb_palette = ["#E69F00","#0072B2"]
# set plot parameters

plot_length= 10 # plot length in inches
plot_width= 10 # plot length in inches
numPC1 = 10 # set number of PC1 intervals
numPC3 = 5 # set number of PC3 intervals
hue = "continent" # select the factor to color by
s = 0.125 # set the scale of the eigenleaves
lf_col = "lightgray" # color of inverse eigenleaf
lf_alpha = 0.5 # alpha of inverse eigenleaf
pt_size = 20 # size of data points
pt_linewidth = 0 # lw of data points, set to 0 for no edges
pt_alpha = 0.6 # alpha of the data points
ax_label_fs = 12 # font size of the x and y axis titles
ax_tick_fs = 8 # font size of the axis ticks
face_col = "white" # color of the plot background
grid_alpha = 0.5 # set the alpha of the grid

plt.figure(figsize=(plot_length, plot_width))

# note that PC2 is asymmetry and always = 0
PC1_vals = np.linspace( np.min(PCs[:,0]), np.max(PCs[:,0]), numPC1 ) # create PC intervals
PC3_vals = np.linspace( np.min(PCs[:,2]), np.max(PCs[:,2]), numPC3 )

for i in PC1_vals: # for each PC1 interval
    for j in PC3_vals: # for each PC2 interval
        
        pc1_val = i # select the current PC1 val
        pc3_val = j # select the current PC3 val

        # calculate the inverse eigenleaf
        inv_leaf = pca.inverse_transform(np.array([pc1_val,0,pc3_val,0,0,0]).reshape(1, -1))
        inv_x = inv_leaf[0,0::2] # select just inverse x vals
        inv_y = inv_leaf[0,1::2] # select just inverse y vals
        
        # plot the inverse eigenleaf
        plt.fill(inv_x*s+pc1_val, inv_y*s+pc3_val, c=lf_col, alpha=lf_alpha)
   
# plot the data on top of the morphospace
sns.scatterplot(data=Newdata, 
                x="PC1", 
                y="PC3", 
                hue=hue, 
                s=pt_size, 
                linewidth=pt_linewidth, 
                alpha=pt_alpha, 
                palette = cb_palette)

plt.legend(bbox_to_anchor=(1.00, 1.02), prop={'size': 8.9})
xlab = "PC1 (49.0%)"
ylab = "PC3 (11.2%)"
plt.xlabel(xlab, fontsize=ax_label_fs)
plt.ylabel(ylab, fontsize=ax_label_fs)
plt.xticks(fontsize=ax_tick_fs)
plt.yticks(fontsize=ax_tick_fs)
plt.gca().set_aspect("equal")
plt.gca().set_facecolor(face_col)
plt.grid(alpha=grid_alpha)
plt.gca().set_axisbelow(True)
plt.show()
#plt.savefig("",bbox_inches='tight')
#Plot edits and merging with PC1 box plots were done in Affinity
```

```{r}
#Apple_metadata<- read.csv("Apple_metadata.csv")
#####Figure 5: PC1/continent boxplot#####
library(ggthemes)
Apple_metadata %>% ggplot(aes(x = continent, y = PC1, fill = continent)) +
  geom_boxplot() +
  theme_few() +
  scale_fill_manual(values = c(
    "North America" = "#0072B2",
    "Asia" = "#E69F00"
  )) +
  labs(fill = "Continent") +
  theme(legend.position = "none") +
  xlab("Continent")

#Saved with small width to fit in plot
#ggsave("Figure5morphospace.pdf",plot=last_plot(), width=2, height=4, units="in")

#Mann-Whitney U test for PC1
wilcox.test(PC1~continent, data=Apple_metadata)
#Plot edits and merging with PC1 box plots were done in Affinity
```


### Figure S3
```{r}
#####Supplemental figure 3######
#Apple_metadata<- read.csv("Apple_metadata.csv")

Apple_metadata %>% ggplot(aes(x = continent, y = PC3, fill = continent)) +
  xlab("Continent") +
  ylab("PC3 (11.2%)")+
  geom_boxplot() +
  theme_few() +
  theme(axis.text.x = element_text(angle = 47, vjust = 1, hjust=1))+
  geom_vline(xintercept=c(4.5))+
  theme(legend.position = "none" ) +
  labs(fill = "Continent") + 
  guides(fill = guide_legend(ncol = 2))  
#ggsave("FigureS3.pdf",plot=last_plot(), width=9.0, height=10.0)

wilcox.test(PC3~continent, data=Apple_metadata)
```

```{r}
#####Supplemental table 3#####
vars_to_test<-c("PC1", "PC3", "area", "width", "length", "aspect", "solidity")
# Run Mann-Whitney U test for each variable
pvals <- sapply(vars_to_test, function(var) {
  wilcox.test(Apple_metadata[[var]] ~ Apple_metadata$continent)$p.value
})


# Adjust p-values for multiple comparisons using bonferroni
pvals_adj <- p.adjust(pvals, method = "bonferroni")  

# Combine results into a data frame
multiresults <- data.frame(
  Variable = vars_to_test,
  Raw_P = pvals,
  Adjusted_P = pvals_adj
)

print(multiresults)

#write.csv(multiresults, "TableS3.csv")

```